Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-19 rnni_convexity.tex
Mareike's comments

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find exact estimates of the radius and diameter of the space, improving the best previously known estimates.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

In this paper, we consider the $\rnni$ space on ranked phylogenetic trees with all taxa being of equal rank.
An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees $\rnniu$.
We postpone accurate definitions until later in this section.

@@

Our definition of a ranked tree implies a natural notion of edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.

@@

In this section we introduce three algorithms, $\findpath$ (Section~\ref{section:alg_findpath}), $\csort$ (Section~\ref{section:alg_csort}), and $\mdtree$ (Section~\ref{section:alg_mdtree}).
$\findpath$ computes a $\rnni$ path between trees.
$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
$\mdtree$ maximises the dissimilarity to find a remote tree given an arbitrary $\rnni$ tree.
While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithms computes a tree with maximum distance if the input tree is a caterpillar tree.
We will prove this fact in Section~\ref{section:diameter}.

@@

Before introducing the algorithm we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and say cluster to mean a non-trivial cluster.
A tree in unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree \autocite{Gavryushkin2018-ol, Gavryushkin2016-uu, Semple2003-nj}.

@@

\proof
It is suffice to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.

Case $k = 1$.
In this case $C_k$ is a cherry $(x, y)$.
The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
Consider the node $w$ preceding $v$ in $T'$, that is, $\rank_{T'}(w) = r - 1$.
If a rank move that swaps $v$ and $w$ is possible then this is the only move on $T'$ that can decrease the rank of $\mrca(x, y)$.
If the rank move is impossible then there is an edge in $T'$ connecting $v$ and $w$.
The only way to decrease the rank of $\mrca(x, y)$ then is to perform an $\nni$ move on that edge.
It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of $\mrca$.

@@

In this section we restrict our attention to the set of caterpillar trees.
These trees are of particular interest because they are used to prove the computing the $\nni$ distance is an NP-hard problem.
As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
Throughout this section we use the list representation of trees described in Section~\ref{section:alg_csort}.

Recall \autocite{Gavryushkin2018-ol} that to compute a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then move the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
This is due to the fact that $\nni$ moves $\rnni$ are only allowed on edges of length one.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, dashed arrow is a move only possible in $\nni$.
Bottom path is a shortest $\rnni$ path.}
\label{fig:NNI_vs_RNNI}
\end{figure}

@@

In this section we discussed the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
Specifically, in $\nni$ there exist two trees $T$ and $R$, which share a split $A | B$, such that every tree on every shortest path between $T$ and $R$ does not have $A | B$.

@@

The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does now hold.

@@

Furthermore, all our methods in this paper for computing shortest paths in special subspaces of $\rnni$ demonstrate that that shared clusters are maintained along shortest paths.

@@

We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
%TODO
Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.

@@

The algorithms we developed and implemented in this paper have served as a main ingredient for our study of the geometry of $\rnni$ space.
These algorithms are of interest on their own as well.
For example, our $\findpath$ algorithm gives a good approximation of the $\rnni$ distance, performing exactly on small trees.
We are not aware of the minimal number of taxa where this algorithm fails to return the correct distance.
The $\mdtree$ algorithm produces trees as far away from a given tree as possible, a task of importance in simulation and model comparison studies.
The $\csort$ algorithm efficiently computes the $\rnni$ distance and shortest paths between caterpillar trees exactly.
To the best of our knowledge no such algorithm exists for $\nni$.
This implies that the algorithmic complexity of comping distances in $\nni$ is higher than in $\rnni$.

@@

We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship awarded to AG.
This work was partially supported by the Ministry of Business, Innovation, and Employment of New Zealand through the Endeavour Smart Ideas and Data Science Programmes grants.

>>>>

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
For example, we find the radius and diameter of the space exactly, improving the best previously known estimates.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we compare geometric and algorithmic properties of the two graphs.
Surprisingly, we discover that both geometric and algorithmic properties of $\rnni$ and $\nni$ are quite different.
For example, we establish convexity of certain natural subspaces in $\rnni$, which are not convex is $\nni$.
Furthermore, our results suggest that the complexity of computing distances in the two graphs is different.
\end{abstract}

@@

In this paper, we consider the $\rnni$ space on ranked phylogenetic trees with all taxa being of equal rank.
An example of such a tree is depicted in Figure~\ref{fig:ranked_tree}.
In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees ($\rnniu$ in their notations).
We postpone accurate definitions until later in this section.

@@

Our definition of a ranked tree implies a natural notion of edge length -- we call $|\rank_T(v) - \rank_T(u)|$ the length of an edge $(u, v)$.
We assume that edges of trees are undirected, so $(u, v) = (v, u)$.

@@

In this section we introduce three algorithms, $\findpath$ (Section~\ref{section:alg_findpath}), $\csort$ (Section~\ref{section:alg_csort}), and $\mdtree$ (Section~\ref{section:alg_mdtree}).
$\findpath$ computes an $\rnni$ path between trees.
$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
$\mdtree$ maximises the dissimilarity to find a remote tree given an arbitrary $\rnni$ tree.
While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.
For example, $\mdtree$ will be an important tool for finding the radius of the $\rnni$ graph, as the algorithms computes a tree with maximum distance if the input tree is a caterpillar tree.
We will prove this fact in Section~\ref{section:diameter}.

@@

Before introducing the algorithm we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
All trees on $n$ leaves share trivial clusters, which are those induced by leaves and the root, so we exclude them from consideration and say cluster to mean a non-trivial cluster.
A tree is unambiguously defined by the list of its clusters sorted according to the rank of the inducing nodes.
We will refer to this list of $n-2$ clusters for a given tree on $n$ taxa as a \emph{cluster representation} of the tree \autocite{Gavryushkin2018-ol, Gavryushkin2016-uu, Semple2003-nj}.

@@

\proof
It suffices to show that the update operation (line~\ref{alg:findpath:line:move_set_down}) is well-defined, that is the $\rnni$ move that decreases the rank of $\mrca_{T'}(C_k)$ is unique.

Case $k = 1$.
In this case $C_k$ is a consists of two taxa $\{x, y\}$.
The node $v = \mrca_{T'}(x, y)$ has $\rank_{T'}(v) = r > 1$.
Consider the node $w$ preceding $v$ in $T'$, that is, $\rank_{T'}(w) = r - 1$.
If a rank move that swaps $v$ and $w$ is possible then this is the only move on $T'$ that can decrease the rank of $\mrca(x, y)$.
If the rank move is impossible then there is an edge in $T'$ connecting $v$ and $w$.
The only way to decrease the rank of $\mrca(x, y)$ then is to perform an $\nni$ move on that edge.
It is not hard to see that out of two possible $\nni$ moves only one decreases the rank of $\mrca$.

@@

In this section we restrict our attention to the set of caterpillar trees.
These trees are of particular interest because they are used to prove that computing the $\nni$ distance is an NP-hard problem.
As we will see below, shortest paths between caterpillar trees in $\rnni$ differ from those in the classical $\nni$ space.
We will show later in this section that $\rnni$ distances between caterpillar trees can be computed in polynomial time.
Throughout this section we use the list representation of trees described in Section~\ref{section:alg_csort}.

Recall \autocite{Gavryushkin2018-ol} that computing a shortest path between two caterpillar trees in $\nni$ sometimes requires first building a clade and then moving the clade around the tree (see Figure~\ref{fig:NNI_vs_RNNI}).
In $\rnni$ however, the necessity of additional rank moves invalidates this strategy.
This is due to the fact that $\nni$ moves in $\rnni$ are only allowed on edges of length one.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{NNI_vs_RNNI}
\vspace{12pt}
\caption{Paths between caterpillar trees $T$ and $R$: Solid arrows indicate paths in $\rnni$, the dashed arrow is a move only possible in $\nni$.
The bottom path is a shortest $\rnni$ path.}
\label{fig:NNI_vs_RNNI}
\end{figure}

@@

In this section we discuss the so-called Split Theorem, which states that every split shared between two trees is maintained along shortest paths between the trees.
For example, Split Theorem is true in the $\spr$ space and false in $\nni$ \autocite{Li1996-zw}.
Specifically, in $\nni$ there exist two trees $T$ and $R$, which share a split $A | B$, such that every tree on every shortest path between $T$ and $R$ does not have $A | B$.

@@

The following example in Figure~\ref{fig:splitthm_counterexample} shows that this form of the Split Theorem does not hold.

@@

Furthermore, all our methods in this paper for computing shortest paths in special subspaces of $\rnni$ demonstrate that shared clusters are maintained along shortest paths.

@@

We have computationally checked that this conjecture is true in $\rnni$ spaces with up to seven taxa ($n = 2, \ldots, 7$).
Specifically, we computed subgraphs of $\rnni$ containing only trees with shared clusters and compared distances in this subgraph with distances in the whole $\rnni$ graph.
We computed these graphs according to the algorithm from \autocite{Gavryushkin2018-ol} as discussed in Section~\ref{section:algorithms}.
The source code of all these implementations can be openly accessed at \autocite{Collienne2019}.

@@

The algorithms we developed and implemented in this paper have served as a main ingredient for our study of the geometry of $\rnni$ space.
These algorithms are of interest on their own as well.
For example, our $\findpath$ algorithm gives a good approximation of the $\rnni$ distance, performing exactly on small trees.
We are not aware of the minimal number of taxa where this algorithm fails to return the correct distance.
The $\mdtree$ algorithm produces trees as far away from a given tree as possible, a task of importance in simulation and model comparison studies.
The $\csort$ algorithm efficiently computes the $\rnni$ distance and shortest paths between caterpillar trees exactly.
To the best of our knowledge no such algorithm exists for $\nni$.
This implies that the algorithmic complexity of computing distances in $\nni$ is higher than in $\rnni$.

@@

We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship (RDF-UOO1702) awarded to AG.
This work was partially supported by the Ministry of Business, Innovation, and Employment of New Zealand through the Endeavour Smart Ideas (CONT-61378-ENDSI-UOO) and Data Science Programmes grants.

MF thanks the joint research project \textit{DIG-IT!} supported by the European Social Fund (ESF), reference: ESF/14-BM-A55-0017/19, and the Ministry of Education, Science, and Culture of Mecklenburg-Vorpommern, Germany, for funding parts of this work.
